---
title: |
  | ![](CHOPMC_logo.pdf){width=5in}  
  | Laurel Redding / CHOPMC-373 / Horses / Shotgun
  | 
  | QC Report
author: "Scott G. Daniel, Ph.D."
date: \today
output: 
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r knitr setup, echo=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=5,
  fig.height=3,
  fig.align = "center"
  )
library(tidyverse)
library(qiimer)
library(vegan)
library(ape)
library(usedist)
library(here)

## Visualization packages
library(pander)
library(gt)

library(pheatbuilder)
library(ggbeeswarm)
library(ggsci)
library(viridis)
library(wesanderson)
library(RColorBrewer)

# stats packages
library(adonisplus)
library(nlme)
library(emmeans) # for lmer post-hoc tests
library(broom)

library(pwr) # for power analysis

library(glue)

panderOptions("knitr.auto.asis", FALSE)

source(here::here("Scripts","Functions","common_functions.R"))
source(here::here("Scripts","Functions","for_functional_analysis.R"))
source(here::here("Scripts","Functions","scott_daniel_useful_R_functions.R"))
```


```{r load}

load(file = here("Data", "preamble_data.RData"))

```


```{r}
study_group_var <- "study_group"
```

\newpage

# Demultiplexing and quality control

## Number of read pairs per sample after demultiplexing

Samples were sequenced on Illumina NovaSeq and demultiplexed. The demultiplexing step involves matching the barcode sequences assicated with each sample to the sequence each read is tagged with.

```{r}
s %>%
  mutate(num_seq=input/1000000) %>%
  ggplot(aes(x=num_seq)) +
    geom_histogram(aes(fill=SampleType), binwidth=0.2) +
    scale_fill_d3(palette = "category20") +
    theme_bw() + 
    theme(
      panel.grid = element_blank()
    ) +
    labs(
      x="Number of read pairs in sample (millions, M)",
      y="Number of samples"
    )
```

## Average nucleotide quality after adapter trimming and quality control

Nextera-XT adapters were removed using trimmomatic-0.33. Nucleotide quality for each position was averaged across all reads using FASTQC.

```{r}
fastqc_df <- read.delim(fastqc_fp, sep='\t')[,1:39] %>%
  pivot_longer(-Samples, names_to="Position", values_to="Quality") %>%
  mutate(
    Position = sub("X", "", Position),
    Position = sub("\\.\\d+", "", Position, perl = TRUE),
    Position = as.numeric(Position)) %>%
  mutate(SampleID=sub("^(.*)_([12])$", "\\1", Samples), Direction=sub("^(.*)_([12])$", "\\2", Samples)) %>%
  mutate(Direction = paste0("R", Direction)) %>%
  mutate(Direction = factor(Direction)) %>%
  filter(SampleID %in% s$SampleID) %>%
  group_by(Direction, Position) %>%
  summarise(MeanQual = mean(Quality, na.rm = T), SdQual = sd(Quality, na.rm = T)) %>%
  ungroup()
fastqc_df %>%
  ggplot(aes(x=Position, y=MeanQual)) + 
    geom_errorbar(aes(ymin=MeanQual - SdQual, ymax=MeanQual + SdQual)) +
    facet_wrap(~ Direction) +
    geom_line() +
    geom_point() +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      strip.background = element_blank()
    ) +
    labs(x='Position in sequence read', 
         y='Average quality score per sample')
```

\newpage

```{r}
host_species <- s %>% filter(!is.na(host_species)) %>% pull(host_species) %>% unique()
```


## Overall distribution of percentage reads removed in quality control

The low quality reads defined by Trimmomatic-0.33 were discarded from further analysis. Human DNA (version GRch38) and Horse DNA (version EquCab3.0) and PhiX DNA were filtered using BWA. Remaining reads were analyzed further.

```{r fig.width=9, fig.height=3}
s %>%
  mutate(`low quality` = (input-host-nonhost) / input) %>%
  mutate(host = host / input) %>%
  mutate(nonhost = nonhost / input) %>%
  filter(!isControl) %>%
  select(SampleID, SampleType, `low quality`, host, nonhost) %>%
  mutate(SampleID = reorder(SampleID, host)) %>%
  pivot_longer(-all_of(c("SampleID", "SampleType")), names_to="category", values_to="percent") %>%
  mutate(category = fct_relevel(category, "low quality", after=0)) %>%
  ggplot(aes(x=SampleID, y=percent, fill=category)) +
    geom_bar(stat="identity") +
    facet_grid(.~SampleType, space="free", scales="free") +
    scale_fill_manual(values=c("#808080", "#A6CEE3", "#1F78B4")) + 
    scale_y_continuous(limits = c(0,1), expand=c(0,0), labels=scales:::percent) +
    theme_classic() +
    theme(
      #axis.text.x=element_text(angle=90,hjust=1,vjust=0.5),
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      strip.background = element_blank()
      ) +
    labs(x="", y="Relative abundance", fill="")
```

## Library concentration and read counts per plate

The samples are processed on a 96-well plate. The figures illustrate the library concentration, read counts after demultiplexing and read counts after quality control / denoising steps for each well. The gray color represents NA values. 

```{r, fig.height=8, fig.width=8}
s %>%
  mutate(plate = paste("Plate", plate)) %>%
  ggplot(aes(x=plate_column, y=fct_rev(plate_row), fill = final_library_conc_ng_ul)) +
    geom_tile(fill = "white") +
    geom_point(shape = 21, size = 10) +
    facet_wrap(~plate, ncol = 1) +
    scale_fill_viridis() +
    scale_x_continuous(breaks=seq(1,12), expand=c(0,0), limit = c(0, 13), position = "top") +
    theme_bw() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      panel.grid = element_blank(),
      aspect.ratio = 0.66
    ) +
    labs(
      x="",
      y="Plate row",
      fill="Library\nconcentration\n(ng/ul)"
    )
```

```{r, fig.height=8, fig.width=8}
s %>%
  mutate(plate = paste("Plate", plate)) %>%
  mutate(input = input/1000000) %>%
  ggplot(aes(x=plate_column, y=fct_rev(plate_row), fill = input)) +
    geom_tile(fill = "white") +
    geom_point(shape = 21, size = 10) +
    facet_wrap(~plate, ncol = 1) +
    scale_fill_viridis() +
    scale_x_continuous(breaks=seq(1,12), expand=c(0,0), limit = c(0, 13), position = "top") +
    theme_bw() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      panel.grid = element_blank(),
      aspect.ratio = 0.66
    ) +
    labs(
      x="",
      y="Plate row",
      fill="Raw read\ncounts,\nMillion (m)"
    )
```


```{r, fig.height=8, fig.width=8}
s %>%
  mutate(plate = paste("Plate", plate)) %>%
  ggplot(aes(x=plate_column, y=fct_rev(plate_row), fill = nonhost)) +
    geom_tile(fill = "white") +
    geom_point(shape = 21, size = 10) +
    facet_wrap(~plate, ncol = 1) +
    scale_fill_viridis() +
    scale_x_continuous(breaks=seq(1,12), expand=c(0,0), limit = c(0, 13), position = "top") +
    theme_bw() +
    theme(
      strip.background = element_blank(),
      strip.placement = "outside",
      panel.grid = element_blank(),
      aspect.ratio = 0.66
    ) +
    labs(
      x="",
      y="Plate row",
      fill="QC read\ncounts,\nMillion (m)"
    )
```

## Taxonomic heatmap

```{r}
prop_cut <- 0.01
```

Taxonomic assignments were performed using the Kraken program.

Heatmap charts were generated from the taxonomic assignments. Each column represents one sample and each row represents one taxon (typically a species). Ranks are included in the plot if the taxon is present in `r 100*prop_cut`% mean abundance in at least one sample type.

The chart is colored white if species were not observed in the sample, dark blue if species were observed at very low abundance.  This allows the reader to quickly survey species presence/absence.  Abundance values exceeding 40% are colored red, indicating an extremely dominant species.


```{r fig.width=20, fig.height=10}
s_toPlot <- s %>%
  mutate(timepoint = ifelse(study_day > 0, "Discharge", "Admit")) %>%
  filter(kraken_counts > 0 | !is.na(kraken_counts)) %>%
  select(SampleID, SampleType, !!enquo(study_group_var), timepoint) %>%
  arrange(timepoint, get(study_group_var)) %>%
  arrange(SampleType) %>%
  droplevels()

order_for_later <- s_toPlot$SampleID

# select taxa with mean relative abundance of 1% in at least one sample type
select_taxa <- summed_props %>%
  as.data.frame() %>% 
  rownames_to_column("Taxa") %>% 
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  right_join(s_toPlot, by="SampleID")  %>%
  group_by(SampleType, Taxa) %>%
  mutate(mean_prop = mean(props)) %>%
  ungroup() %>%
  group_by(Taxa) %>%
  mutate(mean_prop = max(mean_prop)) %>%
  ungroup() %>%
  filter(mean_prop > prop_cut) %>%
  pull(Taxa) %>%
  as.character() %>%
  unique()

props_toPlot <- summed_props[select_taxa, s_toPlot$SampleID]
props_toPlot %>%
  pheat() %>%
  pheat_color_saturated() %>%
  pheat_cluster_rows() %>%
  pheat_annotate_cols(s_toPlot)

```

```{r}

props_toPlot %>%
  pheat() %>%
  pheat_color_saturated() %>%
  pheat_cluster_rows() %>%
  pheat_annotate_cols(s_toPlot) %>%
  pheat_save(filename = here("Figures", "Supplemental_Figure_S1_meat_of_figure.pdf"))

```

```{r second heatmap for replacing samplenames}

library(pheatmap)

clin_md <- read_csv(here("Data", "ClinicalDataFinal.csv")) %>%
  select(-study_group)

s2 <- s %>%
  left_join(clin_md, join_by(subject_id)) %>%
  mutate(timepoint = ifelse(study_day > 0, "Discharge", "Admit"))

s_temp <- s2 %>%
  mutate(timepoint = ifelse(study_day > 0, "Discharge", "Admit")) %>%
  filter(kraken_counts > 0 | !is.na(kraken_counts)) %>%
  select(SampleID, HorseID, SampleType, !!enquo(study_group_var), timepoint) %>%
  mutate(HorseID = ifelse(str_detect(SampleID, "mockdna"), "Positive control", HorseID)) %>%
  mutate(HorseID = ifelse(str_detect(SampleID, "Extract"), "Negative control", HorseID)) %>%
  column_to_rownames(var = "SampleID")
s_temp <- s_temp[order_for_later,]
props_toPlot <- props_toPlot[,order_for_later]
colnames(props_toPlot) <- s_temp$HorseID

fig_name = "Supplmental_Figure_S1_just_for_horseID.pdf"
print(glue("Figure saved as {fig_name}"))

pheatmap(
  props_toPlot,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 9, cellheight = 9, fontsize = 10,
  width = 24, height = 12, annotation_colors = ann_colors, show_colnames = T,
main = "heatmap just for the horseids", filename = here("Figures", fig_name))


```


The chart below is a taxonomic heatmap for the negative and positive control samples. The experimental samples are averaged per sample type. Only the taxa with a mean relative abundance of `r 100*prop_cut`% are shown.

```{r fig.height=10, fig.width=10}
s_toPlot <- s %>%
  filter(kraken_counts > 0 | !is.na(kraken_counts)) %>%
  select(SampleID, SampleType, isControl) %>%
  arrange(SampleType) %>%
  droplevels()

# select taxa with mean relative abundance of 1% in at least one sample type
SampleType_averages <- summed_props %>%
  as.data.frame() %>% 
  rownames_to_column("Taxa") %>% 
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  right_join(s_toPlot, by="SampleID")  %>%
  filter(!isControl) %>%
  group_by(SampleType, Taxa, isControl) %>%
  summarize(props = mean(props)) %>%
  ungroup() %>%
  mutate(SampleID=SampleType)

controls <- summed_props %>%
  as.data.frame() %>% 
  rownames_to_column("Taxa") %>% 
  pivot_longer(-Taxa, names_to="SampleID", values_to="props") %>%
  right_join(s_toPlot, by="SampleID")  %>%
  filter(isControl) 

bind_rows(SampleType_averages, controls) %>%
  group_by(Taxa) %>%
  mutate(mean_prop = mean(props)) %>%
  ungroup() %>%
  filter(mean_prop > prop_cut) %>%
  
  select(SampleID, Taxa, props) %>%
  
  pivot_wider(names_from=SampleID, values_from=props) %>%
  pheat(rownames_in="Taxa") %>%
  pheat_color_saturated() %>%
  pheat_cluster_rows()
```




```{r}
# from now on, exclude control samples
s_toPlot <- s %>%
  filter(!isControl) %>%
  filter(Keep) %>%
  droplevels()
```



## Beta diversity


Here, we use Bray-Curtis and Jaccard distances to compare the species composition of the samples to each other.

The plots below show the distance between each pair of samples in a single 2D plot.  It is not possible to plot the distances exactly on paper, so we have used a method of ordination called Principal Coordinates Analysis to select the best coordinate system for display.  The percentage of total variance captured along each axis is displayed on the chart.

```{r fig.show='hold', out.width='.49\\linewidth', fig.height=3, fig.width=5}
in_both <- intersect(s_toPlot$SampleID, rownames(as.matrix(bc)))

dist_toTest <- dist_subset(bc, in_both)
pc <- pcoa(dist_toTest)
pc_df <- merge(s_toPlot, pc$vectors[, 1:3], by.x="SampleID", by.y="row.names")
pct <- round(pc$values$Relative_eig * 100)
pc_df %>%
  ggplot(aes(x=Axis.1, y=Axis.2, color=host_species)) +
    geom_point() +
    scale_color_d3(palette = "category20") +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(x=paste0("PCoA axis 1 (", pct[1], "%)"), 
         y=paste0("PCoA axis 2 (", pct[2], "%)"),
         color="Sample type", title="Bray-Curtis")



dist_toTest <- dist_subset(jd, in_both)
pc <- pcoa(dist_toTest)
pc_df <- merge(s, pc$vectors[, 1:3], by.x="SampleID", by.y="row.names")
pct <- round(pc$values$Relative_eig * 100)
pc_df %>%
  ggplot(aes(x=Axis.1, y=Axis.2, color=host_species)) +
    geom_point() +
    scale_color_d3(palette = "category20") +
    theme_bw() +
    theme(
      axis.text=element_blank(),
      axis.ticks=element_blank(),
      panel.grid = element_blank(), 
      aspect.ratio = 1,
      strip.background = element_blank()
    ) + 
    labs(x=paste0("PCoA axis 1 (", pct[1], "%)"), 
         y=paste0("PCoA axis 2 (", pct[2], "%)"),
         color="Sample type", title="Jaccard")
```


# Appendix

The table below summarizes the total number of reads per plate after demultiplexing and after quality control / denoising steps.

```{r, results='asis'}
s %>%
  group_by(plate) %>%
  summarize(Demux = sum(input, na.rm=T), QC = sum(nonhost, na.rm = T),  Kraken = sum(kraken_counts, na.rm = T)) %>%
  ungroup() %>%
  pander(split.table=Inf, caption = "Number of reads per plate")
```


The plot below illustrates the number of reads after demultiplexing and the library concentration for each sample. 

```{r fig.height=4, fig.width=5}
s %>%
  ggplot(aes(x=input, y=final_library_conc_ng_ul)) +
    geom_point() +
    theme_bw() +
    labs(
      x="Raw read counts",
      y="Library concentration (ng/ul)"
    )
```


The plot below illustrates the number of reads after demultiplexing and the library concentration for each sample. The x axis is on a log10 scale. 

```{r fig.height=4, fig.width=5}
s %>%
  mutate(input = input + 1) %>%
  ggplot(aes(x=input, y=final_library_conc_ng_ul)) +
    geom_point() +
    scale_x_continuous(trans="log10") +
    theme_bw() +
    labs(
      x="Raw read counts",
      y="Library concentration (ng/ul)"
    )
```


## Number of reads before and after trimmming Illumina adapter sequences with Trimmomatic.

```{r, results='asis'}
s %>%
  arrange(desc(input)) %>%
  select(
    SampleID,
    SampleType,
    `Library concentration ng/ul` = final_library_conc_ng_ul,
    Input = input,
    Dropped = dropped,
    `Forward only` = fwd_only,
    `Reverse only` = rev_only,
    `Both kept` = both_kept) %>%
  pander(split.table = Inf)
```

\newpage

## Number of reads before and after filtering of host genome sequence.

```{r, results='asis'}
s %>%
  mutate(`Percent host reads` = round(100 * host / (host + nonhost),2)) %>%
  arrange(desc(`Percent host reads`)) %>%
  select(
    SampleID,
    `Host reads` = host,
    `Non-host reads` = nonhost,
    `Percent host reads`) %>%
  pander(split.table = Inf)
```

```{r export sample sheet}

s %>%
  write_csv(here("Data", "CHOPMC-373-prepared-sample-sheet.csv"))

```


# End of report

```{r Generate time stamped report, eval=FALSE, include=FALSE}
#notes
#neat: you can run this following command in the console to give your reports custom names (or date-stamp them)
#rmarkdown::render('ronen_stein_Report_shotgun.Rmd',output_file = paste('CEASE.report.', Sys.Date(), '.pdf', sep=''))

rstudioapi::documentSaveAll()

script_dir <- file.path("Scripts","Rmds")
output_dir <- file.path("Output")

report_fp <- here::here(script_dir, "Shotgun_qc_report.Rmd")
output_fp <- here::here(output_dir, paste('CHOPMC-373_qc_report_', Sys.Date(), '.pdf', sep=''))

rmarkdown::render(report_fp, output_file = output_fp)

```

