---
title: |
  ![](CHOPMC_logo.pdf){width=5in}  
  CHOPMC-373_Redding_Shotgun / Horse AMR analysis
author: "Scott G. Daniel, Ph.D."
date: \today
geometry: margin=3cm
Output: 
    pdf_document:
        keep_tex: false
        toc: true
        toc_depth: 5
---

```{r setup, echo=FALSE}
### ================
###   knitr setup
### ================
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=6,
  fig.height=6,
  fig.align = "center",
  out.width = "100%"
  )
```

```{r libraries, message=FALSE, warning=FALSE}
library(here)
library(ggbeeswarm)
library(grid)
library(magrittr)
library(pander)
library(pheatmap)
library(usedist)
library(qiimer)
library(reshape2)
library(jsonlite)
library(vegan)
library(ape)
library(tidyverse)
library(gghighlight)
library(RColorBrewer)
library(patchwork)
library(kylemisc)
library(ggpubr)
library(GGally)
library(glue)
library(wesanderson)

#MOAR cOLORS!!!
library(ggsci)
library(ghibli)
library(qualpalr)
library(emmeans)

panderOptions("knitr.auto.asis", FALSE)
#this also means that you have to put results='asis' in any block that has pander Output_pseudocount
#this lets you loop through variables and produce multiple pander tables and ggplots in a
#single code block!

```

```{r functions}

source(here("Scripts","Functions","common_functions.R"))
source(here("Scripts","Functions","scott_daniel_useful_R_functions.R"))
source(here("Scripts","Functions","for_functional_analysis.R"))

```

```{r load savepoint, include=FALSE}
#you can load data here if you have a savepoint from your main R markdown (but remember that libraries do not get saved)
load(here("Data","after_taxa_analysis.RData"))

table_iterator <- function() {
  table_count <- 0
  function(x, title = "table_export", ...) {
    table_count <<- table_count + 1
    if (table_count < 10) {
      table_count <- paste0("0", table_count)
    }
    table_name <- paste0(title, "_", table_count, ".csv")
    table_path <- here("Output_pseudocount", table_name)
    write_csv(x = x, file = table_path, ...)
    print(paste0("Full table saved as ", table_name, "."))
  }
}

table_export <- table_iterator()
```

# Abx Resistance Genes

Reads were aligned to the Comprehensive Antibiotic Resistance Database (https://card.mcmaster.ca) to assess the antibiotic resistance potential for each sample.

```{r loading CARD}
library(ontologyIndex)
obo <- get_ontology(here("Data","aro.obo"), extract_tags = "everything")
ontology_CARD <- data.frame(ARO = obo$id, name = obo$name) # all CARD ontology, which includes genes and antibiotic drugs
rownames(ontology_CARD) <- NULL

cd_desc <- read_tsv(here("Data","20230525_protein_fasta_protein_homolog_model.txt")) %>%
  select(geneID, ARO) %>%
  unique() %>%
  inner_join(data.frame(ARO = obo$id, description = obo$name), by = "ARO") #just abx resistance ARO descriptions

CARD <- read_table_if_no_rds(here("Data", "sbx_gene_clusters", "20230525_protein_fasta_protein_homolog_model"), name_of_table = CARD, rds_file_name = "CARD.rds") %>%
  left_join(cd_desc, by = "geneID")

gene_lengths <- read_tsv(here("Data","CARD_lengths.tsv"), col_names = c("Description","AA_length")) %>%
  mutate(nt_length = AA_length * 3) %>%
  mutate(ARO = str_extract(Description, "ARO:[0-9]{7}")) %>%
  mutate(ARO = str_trim(ARO)) %>%
  mutate(geneID = str_extract(Description, "(?<=gb\\|)[\\w.]+(?=\\|ARO)")) %>%
  select(ARO, nt_length)

CARD_rpkm <- CARD %>%
  left_join(gene_lengths, by = c("ARO")) %>%
  right_join(s %>% select(SampleID, nonhost), by = "SampleID") %>%
  # group_by(SampleID) %>%
  # mutate(sample_sums = sum(count)) %>% #do it like RNA way
  # ungroup() %>%
  mutate(millr = (nonhost + 1) / 1E6) %>%
  mutate(kbase = nt_length / 1E3) %>%
  mutate(RPKM = (count + 1) / millr / kbase) %>%
  group_by(ARO) %>%
  filter(sum(count) > 0) %>%
  ungroup() %>%
  select(-database)

CARD_annotation <- read_tsv(here("Data", "aro_categories_index.tsv"))

colnames(CARD_annotation) <- c("geneID", "DNA_Accession", "AMR_Gene_Family", "Drug_Class", 
"Resistance_Mechanism")

```

## Admit vs Discharge, stratified by length of Abx

Here, the Discharge will be compared to Admit and stratified by length that the horses were on antibiotics (24hr or 72hr).

```{r, results='asis'}
## Subset the metadata file to only keep the samples you would like to test on
s_toTest <- s2 %>%
  filter(host_species %in% c("Horse")) %>%
  filter(Keep) %>%
  filter(!isControl) %>%
  mutate(study_group = fct_relevel(study_group, "24h", "72h")) %>%
  mutate(timepoint = fct_relevel(timepoint, "Admit")) %>%
  mutate(time_study = paste0(timepoint, study_group)) %>%
  mutate(time_study = fct_relevel(time_study, "Admit24h", "Admit72h", "AdmitMed", "Discharge24h", "Discharge72h")) %>%
  droplevels()

## Set colors for each factor ahead of time so they are consistent through the report.
ann_colors = list(
  time_study = setNames(brewer.pal(length(levels(s_toTest$time_study)), "Set3"), levels(s_toTest$time_study)))
ann_colors = list(time_study = c(Admit24h = "#8DD3C7", Admit72h = "#FFFFB3", 
AdmitMed = "#BEBADA", Discharge24h = "#FB8072", Discharge72h = "#80B1D3", 
DischargeMed = "#FDB462"))

min_percent = 0.99

```

### RPKM analysis

ARO counts from the CARD database were transformed in the following way: counts were divided by the million nonhost reads and the ARO length in kilobases, thus RPKM (Reads per Kilobase per Million nonhost reads). To account for zeros, a pseudocount of unity was added to both the numerator and denominator in that operation. Finally, a logarithm of base 2 was taken for the relative abundance to account for skewed distributions. We will restrict our initial analysis to the top 100 genes that are present (i.e. nonzero before transformation) in at least `r min_percent` of samples.

```{r}

RPKM_toTest <- CARD_rpkm %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  group_by(ARO) %>%
  mutate(perc_present = mean(count > 0)) %>%
  mutate(mean_RPKM = mean(RPKM)) %>%
  ungroup() %>%
  filter(perc_present > min_percent) %>%
  mutate(log2_RPKM = log2(RPKM))

top100 <- RPKM_toTest %>%
  select(ARO, mean_RPKM) %>%
  unique() %>%
  slice_max(order_by = mean_RPKM, n = 100)

RPKM_toTest %<>% filter(ARO %in% top100$ARO)

```

#### Top 100

##### Tests

Linear mixed-effect models were used to estimate the change in log2 of RPKM of selected ARO between time and study_group combined variable. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.05 are shown in the table below. `r length(unique(RPKM_toTest$ARO))` genes were tested. Exported table includes all results.

```{r, results='asis'}

my_form <- "log2_RPKM ~ Age + Sex + time_study"

summaries_df <- run_lmer_on_props(factor_to_test = "ARO", props_toTest = RPKM_toTest, s_toTest = s_toTest, form1 = my_form, p_cutoff = 1, rep_mes_label = "subject_id") %>%
  mutate(term = str_remove(term, "time_study")) %>%
  left_join(cd_desc %>% select(ARO, description), by = "ARO")

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear model with the formula: {my_form}"))
table_export(summaries_df, "gene_table_export_pseudocount")

```

##### Boxplot

```{r fig.height=12, fig.width=8}

to_select <- summaries_df %>%
  filter(fdr < 0.05) %>%
  pull(ARO) %>%
  unique()

number_of_blocks <- ceiling(length(to_select) / 10)
chunk_it <- function(x,n) split(x, cut(seq_along(x), n, labels = FALSE))

if (number_of_blocks > 1) {
  my_blocks <- chunk_it(to_select, number_of_blocks)
} else {
  my_blocks <- list(to_select)
}

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]

my_plot <- RPKM_toTest %>%
  right_join(s_toTest, by = "SampleID") %>%
  filter(ARO %in% subset) %>%
  mutate(label = paste(ARO, description, sep = "\n")) %>%
  mutate(label = reorder(label, -RPKM)) %>%
  ggplot(aes(x=time_study, y=RPKM, color=time_study)) +
  geom_boxplot(outlier.alpha=0) +
  geom_quasirandom() +
  facet_wrap(~label, scales="free", ncol = 5) +
  scale_color_manual(values=ann_colors$time_study) +
  scale_y_continuous(trans = "log10") +
  theme_bw() +
  theme(
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    panel.grid = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    y="RPKM",
    x=""
  )

plot(my_plot)

ggsave(here("Output_pseudocount", glue("horses_ARO_{i}.pdf")), height = 6, width = 8)

}

```

##### Post-hoc multiple comparisons

Since we are also interested in other time_study-to-time_study comparisons, we will do post-hoc multiple comparisons with estimated marginal means. Exported table includes all results.

```{r, results='asis'}

summaries_df <- RPKM_toTest %>%
  inner_join(s_toTest, by = "SampleID") %>%
  group_by(ARO) %>%
  do(broom::tidy(emmeans(object = nlme::lme(log2_RPKM ~ Age + Sex + time_study, data=., na.action=na.omit, random = ~1|subject_id), specs = tukey ~ time_study, reverse = T)$contrasts)) %>%
  ungroup() %>%
  select(-term, -null.value) %>%
  left_join(cd_desc %>% select(ARO, description), by = "ARO")

save(summaries_df, file = here("Data", "100_ARO_posthoc_tests_horse_pseudocount.RDS"))

load(file = here("Data", "100_ARO_posthoc_tests_horse_pseudocount.RDS"))

summaries_df %>%
  filter(adj.p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = paste(glue("Estimated marginal means pairwise statistical tests on previous linear model. P-value is adjusted with Tukey's method and filtered by < 0.05.")))

table_export(x = summaries_df, title = "gene_table_export_pseudocount")

```

##### Heatmap

We will make a heatmap. We will filter to ARO's that were significant in at least one comparison. We will scale the RPKM's per ARO to better compare between ARO's.

```{r, fig.width=30, fig.height=50, out.width="80%"}

just_sig <- summaries_df %>%
  filter(adj.p.value < 0.05)

my_AROs <- summaries_df %>%
  pivot_wider(id_cols = ARO, names_from = contrast, values_from = adj.p.value) %>%
  filter(ARO %in% just_sig$ARO)

CARD_subset <- CARD_rpkm %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  filter(ARO %in% my_AROs$ARO)

ARO_z_score <- CARD_subset %>%
  group_by(ARO) %>%
  mutate(ARO_zscore = scale(RPKM)) %>%
  ungroup()

ARO_heatmap <- ARO_z_score %>%
  spread_to_numeric_matrix(row_key = "description", col_key = "SampleID", value = "ARO_zscore")

group_by <- c("time_study")
grps <- c(group_by)
grps_symbol <- syms(grps)
s_toTest %<>% arrange(!!!grps_symbol)

anno <- s_toTest[,grps] %>% as.data.frame()
rownames(anno) <- s_toTest$SampleID
colnames(anno) <- grps

ARO_heatmap <- ARO_heatmap[,rownames(anno)] #sorts the heatmap according to the annotation

paletteLength <- 50
myColor <- colorRampPalette(c("yellow", "white", "blue"))(paletteLength)
myBreaks <- c(seq(min(ARO_heatmap), 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(max(ARO_heatmap)/paletteLength, max(ARO_heatmap), length.out=floor(paletteLength/2)))

pheatmap(
  ARO_heatmap,
  annotation = anno,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 9, cellheight = 9, fontsize = 10,
  width = 24, height = 12, annotation_colors = ann_colors, show_colnames = T,
main = "Z-score of AMR genes in horses")

fig_name = "sig_changed_ARO_zscore_heatmap_horses.pdf"
print(glue("Figure saved as {fig_name}"))

pheatmap(
  ARO_heatmap,
  annotation = anno,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 12, cellheight = 12, fontsize = 12,
  width = 24, height = 12, annotation_colors = ann_colors, show_colnames = T,
main = "Z-score of AMR genes in horses", filename = here("Output_pseudocount", fig_name))

```

##### Richness by Study group

Here, we will plot the unique number of ARO's across study groups. And create a table.

```{r}

CARD_ARO_richness <- CARD %>%
  inner_join(s_toTest) %>%
  filter(count > 0) %>%
  select(SampleID, ARO) %>%
  unique() %>%
  group_by(SampleID) %>%
  summarise(Richness_ARO = n())

my_plot <- CARD_ARO_richness %>%
  inner_join(s_toTest, by = "SampleID") %>%
  ggplot(aes(x = time_study, y = Richness_ARO)) + 
  geom_boxplot(outlier.alpha = 0) +
  geom_quasirandom() +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust = 1)) +
  xlab(label = "Zinc interact") + ylab("Average Richness (number of unique AROs)") +
  ggtitle(label = "Richness")

my_plot

ggsave(here("Output_pseudocount", "richness of AROs horses.pdf"), height = 6, width = 6)

my_table <- CARD_ARO_richness %>%
  inner_join(s_toTest, by = "SampleID") %>%
  group_by(time_study) %>%
  summarise(mean_richness = mean(Richness_ARO), sd_richness = sd(Richness_ARO), n = n())

my_table %>%
  write_csv(here("Output_pseudocount", "Richness of AROs per study group.csv"))

```

##### Statistical test

Might as well do a statistical test.

```{r, results='asis'}
my_lm <- CARD_ARO_richness %>%
  inner_join(s_toTest, by = "SampleID") %>%
  nlme::lme(Richness_ARO ~ Age + Sex + time_study, data = ., random = ~1|subject_id)

summaries_df <- broom::tidy(emmeans(object = my_lm, specs = tukey ~ time_study, reverse = T)$contrasts) %>%
  select(-term, -null.value)

table_export(summaries_df, "gene_table_export_pseudocount")

summaries_df %>%
  pander(split.cells = Inf, digits = 2)


```

#### 50% Threshold

Here we will filter to ARO's that are nonzero in at least 50% of samples.

```{r}

min_percent <- 0.5

RPKM_toTest <- CARD_rpkm %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  group_by(ARO) %>%
  mutate(perc_present = mean(count > 0)) %>%
  mutate(mean_RPKM = mean(RPKM)) %>%
  ungroup() %>%
  filter(perc_present > min_percent) %>%
  group_by(SampleID) %>%
  mutate(rel_RPKM = (RPKM + 1) / (sum(RPKM) + 1) ) %>%
  ungroup() %>%
  mutate(log2_RPKM = log2(rel_RPKM))

```

##### Tests

Linear mixed-effect models were used to estimate the change in RPKM of selected ARO between study groups. The RPKMs were log2 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.05 are shown in the table below. `r length(unique(RPKM_toTest$ARO))` genes were tested. Exported table includes all results.

```{r, results='asis'}

my_form <- "log2_RPKM ~ Age + Sex + time_study"

summaries_df <- run_lmer_on_props(factor_to_test = "ARO", props_toTest = RPKM_toTest, s_toTest = s_toTest, form1 = my_form, p_cutoff = 1, rep_mes_label = "subject_id") %>%
  mutate(term = str_remove(term, "time_study")) %>%
  left_join(cd_desc %>% select(ARO, description), by = "ARO")

# summaries_df %>%
#   filter(p.value < 0.05) %>%
#   pander(split.table=Inf, digits=2, caption = glue("Linear model with the formula: {my_form}"))

table_export(summaries_df, "gene_table_export_pseudocount")

```

##### Boxplot

See individual pdfs.

```{r fig.height=12, fig.width=8}

to_select <- summaries_df %>%
  filter(fdr < 0.05) %>%
  pull(ARO) %>%
  unique()

number_of_blocks <- ceiling(length(to_select) / 10)
chunk_it <- function(x,n) split(x, cut(seq_along(x), n, labels = FALSE))

if (number_of_blocks > 1) {
  my_blocks <- chunk_it(to_select, number_of_blocks)
} else {
  my_blocks <- list(to_select)
}

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]

my_plot <- RPKM_toTest %>%
  right_join(s_toTest, by = "SampleID") %>%
  filter(ARO %in% subset) %>%
  mutate(label = paste(ARO, description, sep = "\n")) %>%
  mutate(label = reorder(label, -RPKM)) %>%
  ggplot(aes(x=time_study, y=RPKM, color=time_study)) +
  geom_boxplot(outlier.alpha=0) +
  geom_quasirandom() +
  facet_wrap(~label, scales="free", ncol = 5) +
  scale_color_manual(values=ann_colors$time_study) +
  scale_y_continuous(trans = "log10") +
  theme_bw() +
  theme(
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    panel.grid = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    y="RPKM",
    x=""
  )

ggsave(here("Output_pseudocount", glue("horses_ARO_50perc_thresh_{i}.pdf")), height = 6, width = 8)

}

```

##### Post-hoc multiple comparisons

Since we are also interested in other group-to-group comparisons, we will do post-hoc multiple comparisons with estimated marginal means. Exported table includes all results.

```{r, results='asis'}

summaries_df <- RPKM_toTest %>%
  inner_join(s_toTest, by = "SampleID") %>%
  group_by(ARO) %>%
  do(broom::tidy(emmeans(object = nlme::lme(log2_RPKM ~ Age + Sex + time_study, data=., na.action=na.omit, random = ~1|subject_id), specs = tukey ~ time_study, reverse = T)$contrasts)) %>%
  ungroup() %>%
  select(-term, -null.value) %>%
  left_join(cd_desc %>% select(ARO, description), by = "ARO")

save(summaries_df, file = here("Data", "50perc_tresh_posthoc_tests_horse_pseudocount.RDS"))

load(file = here("Data", "50perc_tresh_posthoc_tests_horse_pseudocount.RDS"))

# summaries_df %>%
#   filter(adj.p.value < 0.05) %>%
#   pander(split.table=Inf, digits=2, caption = paste(glue("Estimated marginal means pairwise statistical tests on previous linear model. P-value is adjusted with Tukey's method and filtered by < 0.05.")))

table_export(x = summaries_df, title = "gene_table_export_pseudocount")

```

##### Heatmap

```{r}
sig_cutoff <- 0.01
```


We will make a heatmap. We will filter to ARO's that were significant (adj.p.value < `sig_cutoff`) in at least one comparison. We will scale the RPKM's per ARO to better compare between ARO's.

```{r, fig.width=30, fig.height=50, out.width="80%"}

just_sig <- summaries_df %>%
  filter(adj.p.value < sig_cutoff)

my_AROs <- summaries_df %>%
  pivot_wider(id_cols = ARO, names_from = contrast, values_from = adj.p.value) %>%
  filter(ARO %in% just_sig$ARO)

CARD_subset <- CARD_rpkm %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  filter(ARO %in% my_AROs$ARO)

ARO_z_score <- CARD_subset %>%
  group_by(ARO) %>%
  mutate(ARO_zscore = scale(RPKM)) %>%
  ungroup()

ARO_heatmap <- ARO_z_score %>%
  spread_to_numeric_matrix(row_key = "description", col_key = "SampleID", value = "ARO_zscore")

group_by <- c("time_study")
grps <- c(group_by)
grps_symbol <- syms(grps)
s_toTest %<>% arrange(!!!grps_symbol)

anno <- s_toTest[,grps] %>% as.data.frame()
rownames(anno) <- s_toTest$SampleID
colnames(anno) <- grps

ARO_heatmap <- ARO_heatmap[,rownames(anno)] #sorts the heatmap according to the annotation

paletteLength <- 50
myColor <- colorRampPalette(c("yellow", "white", "blue"))(paletteLength)
myBreaks <- c(seq(min(ARO_heatmap), 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(max(ARO_heatmap)/paletteLength, max(ARO_heatmap), length.out=floor(paletteLength/2)))

pheatmap(
  ARO_heatmap,
  annotation = anno,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 9, cellheight = 9, fontsize = 10,
  width = 24, height = 12, annotation_colors = ann_colors, show_colnames = T,
main = "Z-score of AMR genes in horses")

fig_name = "sig_changed_ARO_zscore_heatmap_horses_50perc_thresh.pdf"
print(glue("Figure saved as {fig_name}"))

pheatmap(
  ARO_heatmap,
  annotation = anno,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 12, cellheight = 12, fontsize = 12,
  width = 24, height = 12, annotation_colors = ann_colors, show_colnames = T,
main = "Z-score of AMR genes in horses", filename = here("Output_pseudocount", fig_name))

```


#### Group by AMR Gene Family

Here, we will first time_study ARO genes by their AMR Gene Families and then sum up the RPKM per family. We will look at the top 100 by mean RPKM.

```{r}

AMR100 <- CARD_rpkm %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  left_join(CARD_annotation, by = "geneID") %>%
  group_by(SampleID, AMR_Gene_Family) %>%
  summarise(sum_rpkm = sum(RPKM)) %>%
  group_by(AMR_Gene_Family) %>%
  summarise(x = mean(sum_rpkm)) %>%
  slice_max(order_by = x, n = 100)

CARD_AMR <- CARD_rpkm %>%
  left_join(CARD_annotation, by = "geneID") %>%
  group_by(SampleID, AMR_Gene_Family) %>%
  summarise(sum_rpkm = sum(RPKM)) %>%
  filter(AMR_Gene_Family %in% AMR100$AMR_Gene_Family) %>%
  mutate(log2_sum_rpkm = log2(sum_rpkm))

```

##### Tests

Linear mixed-effect models were used to estimate the change in RPKM of selected AMR gene family between study time_studys. The RPKMs were log2 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.05 are shown in the table below.

```{r, results='asis'}

my_form <- "log2_sum_rpkm ~ Age + Sex + time_study"

temp_df <- tibble()

for (selected_gene in unique(CARD_AMR$AMR_Gene_Family)) {

  props_toTest <- CARD_AMR %>%
  filter(AMR_Gene_Family %in% selected_gene) %>%
  inner_join(s_toTest, by = "SampleID")

  possible_error <- try(tiny_df <- tidy_lmer(nlme::lme(as.formula(my_form), random = as.formula("~ 1 | subject_id"), data=props_toTest, na.action=na.omit)) %>% filter(term != '(Intercept)'))

  tiny_df$AMR_Gene_Family <- selected_gene

  if (class(possible_error)[1] %in% "try-error") {
    print(glue("Error with {selected_gene}"))
    tiny_df <- tibble()
  }

  temp_df <- bind_rows(temp_df, tiny_df)

}

summaries_df <- temp_df %>%
  group_by(term) %>%
  mutate(fdr = p.adjust(p.value, method="BH")) %>%
  ungroup() %>%
  select(AMR_Gene_Family, everything())

summaries_df %>%
  filter(fdr < 0.05) %>%
  mutate(term = str_remove(term, "time_study")) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear model with the formula: {my_form}"))

table_export(summaries_df, "gene_table_export_pseudocount")
```

##### Boxplot

```{r fig.height=12, fig.width=8}

to_select <- summaries_df %>%
  filter(fdr < 0.05) %>%
  pull(AMR_Gene_Family) %>%
  unique()

number_of_blocks <- ceiling(length(to_select) / 7)
my_blocks <- chunk_it(to_select, number_of_blocks)

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]

my_plot <- CARD_AMR %>%
  right_join(s_toTest, by = "SampleID") %>%
  filter(AMR_Gene_Family %in% subset) %>%
  mutate(AMR_Gene_Family = reorder(AMR_Gene_Family, -sum_rpkm)) %>%
  ggplot(aes(x=time_study, y=sum_rpkm, color=time_study)) +
  geom_boxplot(outlier.alpha=0) +
  geom_quasirandom() +
  facet_wrap(~AMR_Gene_Family, scales="free", ncol = 4, strip.position = "right") +
  scale_color_manual(values=ann_colors$time_study) +
  scale_y_continuous(trans = "log10") +
  theme_bw() +
  theme(
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    panel.grid = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    y="sum_RPKM",
    x=""
  )

plot(my_plot)

ggsave(here("Output_pseudocount", glue("horses_AMRgenefamily_{i}.pdf")), height = 8, width = 6)

}

```

##### Post-hoc multiple comparisons

Since we are also interested in other time_study-to-time_study comparisons, we will do post-hoc multiple comparisons with estimated marginal means. Exported table includes all results.

```{r, results='asis'}

#filtering out VHH beta-lactamase because its the same in every group
#and VHW beta-lactamase because it throws an error

summaries_df <- CARD_AMR %>%
  filter(!AMR_Gene_Family %in% c("VHW beta-lactamase", "VHH beta-lactamase")) %>%
  inner_join(s_toTest, by = "SampleID") %>%
  group_by(AMR_Gene_Family) %>%
  do(broom::tidy(emmeans(object = nlme::lme(log2_sum_rpkm ~ Age + Sex + time_study, data=., na.action=na.omit, random = ~1|subject_id), specs = tukey ~ time_study, reverse = T)$contrasts)) %>%
  ungroup() %>%
  select(-term, -null.value)

save(summaries_df, file = here("Data", "top100_AMRgenefamily_posthoc_tests_horse_pseudocount.RDS"))

load(file = here("Data", "top100_AMRgenefamily_posthoc_tests_horse_pseudocount.RDS"))

summaries_df %>%
  filter(adj.p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = paste(glue("Estimated marginal means pairwise statistical tests on previous linear model. P-value is adjusted with Tukey's method and filtered by < 0.05.")))

table_export(x = summaries_df, title = "gene_table_export_pseudocount")

```

##### Heatmap

```{r}
sig_cutoff <- 0.05
```

Heatmap of these AMR gene families with adj.p.value cutoff of `sig_cutoff`.

```{r, fig.width=30, fig.height=50, out.width="80%"}

just_sig <- summaries_df %>%
  filter(adj.p.value < sig_cutoff)

CARD_AMR_subset <- CARD_AMR %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  filter(AMR_Gene_Family %in% just_sig$AMR_Gene_Family)

AMR_z_score <- CARD_AMR_subset %>%
  group_by(AMR_Gene_Family) %>%
  mutate(zscore = scale(sum_rpkm)) %>%
  ungroup()

AMR_heatmap <- AMR_z_score %>%
  spread_to_numeric_matrix(row_key = "AMR_Gene_Family", col_key = "SampleID", value = "zscore")

group_by <- c("time_study")
grps <- c(group_by)
grps_symbol <- syms(grps)
s_toTest %<>% arrange(!!!grps_symbol)

anno <- s_toTest[,grps] %>% as.data.frame()
rownames(anno) <- s_toTest$SampleID
colnames(anno) <- grps

AMR_heatmap <- AMR_heatmap[,rownames(anno)] #sorts the heatmap according to the annotation
order_for_later <- rownames(anno)

paletteLength <- 50
myColor <- colorRampPalette(c("yellow", "white", "blue"))(paletteLength)
myBreaks <- c(seq(min(AMR_heatmap), 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(max(AMR_heatmap)/paletteLength, max(AMR_heatmap), length.out=floor(paletteLength/2)))

pheatmap(
  AMR_heatmap,
  annotation = anno,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 9, cellheight = 9, fontsize = 10,
  width = 24, height = 12, annotation_colors = ann_colors, show_colnames = T,
main = "Z-score by AMR gene family in horses")

fig_name = "sig_changed_AMR_gene_family_zscore_heatmap_horses.pdf"
print(glue("Figure saved as {fig_name}"))

pheatmap(
  AMR_heatmap,
  annotation = anno,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 12, cellheight = 12, fontsize = 12,
  width = 24, height = 12, annotation_colors = ann_colors, show_colnames = T,
main = "Z-score by AMR gene family in horses", filename = here("Output_pseudocount", fig_name))

```

##### Second heatmap for figure construction

```{r, fig.width=30, fig.height=50, out.width="80%"}

just_sig <- summaries_df %>%
  filter(adj.p.value < sig_cutoff)

CARD_AMR_subset <- CARD_AMR %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  filter(AMR_Gene_Family %in% just_sig$AMR_Gene_Family)

AMR_z_score <- CARD_AMR_subset %>%
  group_by(AMR_Gene_Family) %>%
  mutate(zscore = scale(sum_rpkm)) %>%
  ungroup()

AMR_heatmap <- AMR_z_score %>%
  spread_to_numeric_matrix(row_key = "AMR_Gene_Family", col_key = "SampleID", value = "zscore")

s_temp <- s_toTest %>%
  column_to_rownames(var = "SampleID")
s_temp <- s_temp[order_for_later,]
AMR_heatmap <- AMR_heatmap[,order_for_later]
colnames(AMR_heatmap) <- s_temp$HorseID

paletteLength <- 50
myColor <- colorRampPalette(c("yellow", "white", "blue"))(paletteLength)
myBreaks <- c(seq(min(AMR_heatmap), 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(max(AMR_heatmap)/paletteLength, max(AMR_heatmap), length.out=floor(paletteLength/2)))

pheatmap(
  AMR_heatmap,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 9, cellheight = 9, fontsize = 10,
  width = 24, height = 12, show_colnames = T,
main = "Z-score by AMR gene family in horses")

fig_name = "sig_changed_AMR_gene_family_zscore_heatmap_horses_A.pdf"
print(glue("Figure saved as {fig_name}"))

pheatmap(
  AMR_heatmap,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 12, cellheight = 12, fontsize = 12,
  width = 24, height = 12, show_colnames = T,
main = "Z-score by AMR gene family in horses", filename = here("Output_pseudocount", fig_name))

```

#### Group by Drug Class

Here, we will first group ARO genes by their Drug Class and then sum up the RPKM per class. We will look at the top 100 by mean RPKM.

```{r}

DC100 <- CARD_rpkm %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  left_join(CARD_annotation, by = "geneID") %>%
  filter(!Drug_Class %in% "N/A") %>%
  group_by(SampleID, Drug_Class) %>%
  summarise(sum_rpkm = sum(RPKM)) %>%
  group_by(Drug_Class) %>%
  summarise(x = mean(sum_rpkm)) %>%
  slice_max(order_by = x, n = 100)

CARD_DC <- CARD_rpkm %>%
  left_join(CARD_annotation, by = "geneID") %>%
  group_by(SampleID, Drug_Class) %>%
  summarise(sum_rpkm = sum(RPKM)) %>%
  filter(Drug_Class %in% DC100$Drug_Class) %>%
  mutate(log2_sum_rpkm = log2(sum_rpkm))

```

##### Tests

Linear mixed-effect models were used to estimate the change in RPKM of selected AMR gene family between study time_studys. The RPKMs were log2 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.05 are shown in the table below.

```{r, results='asis'}

my_form <- "log2_sum_rpkm ~ Age + Sex + time_study"

summaries_df <- run_lmer_on_props(factor_to_test = "Drug_Class", props_toTest = CARD_DC, s_toTest = s_toTest, form1 = my_form, p_cutoff = 1, rep_mes_label = "subject_id") %>%
  mutate(term = str_remove(term, "time_study"))

summaries_df %>%
  filter(fdr < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear model with the formula: {my_form}"))
table_export(summaries_df, "gene_table_export_pseudocount")
```

##### Boxplot

```{r fig.height=12, fig.width=8}

to_select <- summaries_df %>%
  filter(fdr < 0.05) %>%
  pull(Drug_Class) %>%
  unique()

my_plot <- CARD_DC %>%
  right_join(s_toTest, by = "SampleID") %>%
  filter(Drug_Class %in% to_select) %>%
  mutate(Drug_Class = reorder(Drug_Class, -sum_rpkm)) %>%
  ggplot(aes(x=time_study, y=sum_rpkm, color=time_study)) +
  geom_boxplot(outlier.alpha=0) +
  geom_quasirandom() +
  facet_wrap(~Drug_Class, scales="free", ncol = 4, strip.position = "right") +
  scale_color_manual(values=ann_colors$time_study) +
  scale_y_continuous(trans = "log10") +
  theme_bw() +
  theme(
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    panel.grid = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    y="sum_RPKM",
    x=""
  )

plot(my_plot)

ggsave(here("Output_pseudocount", glue("horses_ARO_drugclass.pdf")), height = 8, width = 6)


```

##### Richness by Drug Class

Here, we will plot the unique number of ARO's per drug class across study groups. We will filter to the top 19 that changed in the linear models.

```{r}

top19 <- summaries_df %>%
  group_by(term) %>%
  slice_min(fdr, n = 19) %>%
  ungroup() %>%
  slice_min(fdr, n = 25) %>%
  select(Drug_Class) %>%
  unique()

CARD_DC_richness <- CARD %>%
  left_join(CARD_annotation, by = "geneID") %>%
  inner_join(s_toTest) %>%
  filter(count > 0) %>%
  select(time_study, Drug_Class, ARO) %>%
  unique() %>%
  group_by(time_study, Drug_Class) %>%
  summarise(Richness = n()) %>%
  filter(!Drug_Class %in% "N/A") %>%
  filter(Drug_Class %in% top19$Drug_Class) #%>%
  # mutate(Drug_Class = str_trunc(Drug_Class, 30))

my_plot <- ggplot(CARD_DC_richness, aes(fill=Drug_Class, x = time_study, y = Richness)) + 
  geom_bar(stat="identity", position = "stack", width = 0.8) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust = 1)) +
  scale_y_continuous(expand = c(0, 0)) +
  xlab(label = "time_study") + ylab("Average Richness (number of unique AMRs)") +
  viridis::scale_fill_viridis(option = "turbo", discrete = T, begin = 0, end = 1, name="Drug class", labels = function(x) str_wrap(x, width = 30)) + # rainbow
  #scale_fill_igv() +
  labs(x = "")

my_plot

ggsave(here("Output_pseudocount", "richness of drug classes horses.pdf"), height = 12, width = 12)

CARD %>%
  left_join(CARD_annotation, by = "geneID") %>%
  inner_join(s_toTest) %>%
  filter(count > 0) %>%
  select(time_study, Drug_Class, ARO) %>%
  unique() %>%
  group_by(time_study, Drug_Class) %>%
  summarise(Richness = n()) %>%
  filter(!Drug_Class %in% "N/A") %>%
  write_csv(here("Output_pseudocount", "Richness of drug classes per study time_study.csv"))

```

##### Post-hoc multiple comparisons

Since we are also interested in other time_study-to-time_study comparisons, we will do post-hoc multiple comparisons with estimated marginal means. Exported table includes all results.

```{r, results='asis'}

summaries_df <- CARD_DC %>%
  inner_join(s_toTest, by = "SampleID") %>%
  group_by(Drug_Class) %>%
  do(broom::tidy(emmeans(object = nlme::lme(log2_sum_rpkm ~ Age + Sex + time_study, data=., na.action=na.omit, random = ~1|subject_id), specs = tukey ~ time_study, reverse = T)$contrasts)) %>%
  ungroup() %>%
  select(-term, -null.value)

save(summaries_df, file = here("Data", "top100_Drug_Class_posthoc_tests_horse_pseudocount.RDS"))

load(file = here("Data", "top100_Drug_Class_posthoc_tests_horse_pseudocount.RDS"))

summaries_df %>%
  filter(adj.p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = paste(glue("Estimated marginal means pairwise statistical tests on previous linear model. P-value is adjusted with Tukey's method and filtered by < 0.05.")))

table_export(x = summaries_df, title = "gene_table_export_pseudocount")

```

# Filtering to AMR's in penem drug class

We will still want to filter to genes that are present in at least 10% of the samples.

```{r}

RPKM_toTest <- CARD_rpkm %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  left_join(CARD_annotation, by = "geneID") %>%
  filter(Drug_Class %in% "cephalosporin;monobactam;penam;penem") %>%
  group_by(ARO) %>%
  mutate(perc_present = mean(count > 0)) %>%
  mutate(mean_RPKM = mean(RPKM)) %>%
  ungroup() %>%
  filter(perc_present > 0.1) %>%
  mutate(log2_RPKM = log2(RPKM))

```

Linear mixed-effect models were used to estimate the change in RPKM of selected ARO between study time_studys. The RPKMs were log2 transformed. Multiple tests were adjusted for false discovery rate using Benjamini-Hochberg method. Only the terms with p<0.05 are shown in the table below. `r length(unique(RPKM_toTest$ARO))` genes were tested. Exported table includes all results.

```{r, results='asis'}

my_form <- "log2_RPKM ~ Age + Sex + time_study"

summaries_df <- run_lmer_on_props(factor_to_test = "ARO", props_toTest = RPKM_toTest, s_toTest = s_toTest, form1 = my_form, p_cutoff = 1, rep_mes_label = "subject_id") %>%
  mutate(term = str_remove(term, "time_study"))

summaries_df %>%
  filter(p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = glue("Linear model with the formula: {my_form}"))
table_export(summaries_df, "gene_table_export_pseudocount")

```

## Boxplot

```{r fig.height=12, fig.width=8}

to_select <- summaries_df %>%
  filter(fdr < 0.05) %>%
  pull(ARO) %>%
  unique()

number_of_blocks <- ceiling(length(to_select) / 10)
chunk_it <- function(x,n) split(x, cut(seq_along(x), n, labels = FALSE))

if (number_of_blocks > 1) {
  my_blocks <- chunk_it(to_select, number_of_blocks)
} else {
  my_blocks <- list(to_select)
}

for (i in 1:number_of_blocks) {
  subset <- my_blocks[[i]]

my_plot <- RPKM_toTest %>%
  right_join(s_toTest, by = "SampleID") %>%
  filter(ARO %in% subset) %>%
  mutate(label = paste(ARO, description, sep = "\n")) %>%
  mutate(label = reorder(label, -RPKM)) %>%
  ggplot(aes(x=time_study, y=RPKM, color=time_study)) +
  geom_boxplot(outlier.alpha=0) +
  geom_quasirandom() +
  facet_wrap(~label, scales="free", ncol = 5) +
  scale_color_manual(values=ann_colors$time_study) +
  scale_y_continuous(trans = "log10") +
  theme_bw() +
  theme(
    axis.text.x=element_text(angle=90, hjust=1, vjust=0.5),
    panel.grid = element_blank(),
    strip.background = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    y="RPKM",
    x=""
  )

plot(my_plot)

ggsave(here("Output_pseudocount", glue("horses_penem_ARO_{i}.pdf")), height = 6, width = 8)

}

```

## Post-hoc multiple comparisons

Since we are also interested in other time_study-to-time_study comparisons, we will do post-hoc multiple comparisons with estimated marginal means. Exported table includes all results.

```{r, results='asis'}

summaries_df <- RPKM_toTest %>%
  inner_join(s_toTest, by = "SampleID") %>%
  group_by(ARO) %>%
  do(broom::tidy(emmeans(object = nlme::lme(log2_RPKM ~ Age + Sex + time_study, data=., na.action=na.omit, random = ~1|subject_id), specs = tukey ~ time_study, reverse = T)$contrasts)) %>%
  ungroup() %>%
  select(-term, -null.value) %>%
  left_join(cd_desc %>% select(ARO, description), by = "ARO")

summaries_df %>%
  # filter(adj.p.value < 0.05) %>%
  pander(split.table=Inf, digits=2, caption = paste(glue("Estimated marginal means pairwise statistical tests on previous linear model. P-value is adjusted with Tukey's method and filtered by < 0.05.")))

table_export(x = summaries_df, title = "gene_table_export_pseudocount")

```

## Heatmap

We will make a heatmap. We will filter to ARO's that were significant in at least one comparison. We will scale the RPKM's per ARO to better compare between ARO's.

```{r, fig.width=30, fig.height=50, out.width="80%"}

just_sig <- summaries_df %>%
  filter(adj.p.value < 0.05)

my_AROs <- summaries_df %>%
  pivot_wider(id_cols = ARO, names_from = contrast, values_from = adj.p.value) %>%
  filter(ARO %in% just_sig$ARO)

CARD_subset <- CARD_rpkm %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  filter(ARO %in% my_AROs$ARO)

ARO_z_score <- CARD_subset %>%
  group_by(ARO) %>%
  mutate(ARO_zscore = scale(RPKM)) %>%
  ungroup()

ARO_heatmap <- ARO_z_score %>%
  spread_to_numeric_matrix(row_key = "description", col_key = "SampleID", value = "ARO_zscore")

group_by <- c("time_study")
grps <- c(group_by)
grps_symbol <- syms(grps)
s_toTest %<>% arrange(!!!grps_symbol)

anno <- s_toTest[,grps] %>% as.data.frame()
rownames(anno) <- s_toTest$SampleID
colnames(anno) <- grps

ARO_heatmap <- ARO_heatmap[,rownames(anno)] #sorts the heatmap according to the annotation
order_for_later <- rownames(anno)

paletteLength <- 50
myColor <- colorRampPalette(c("yellow", "white", "blue"))(paletteLength)
myBreaks <- c(seq(min(ARO_heatmap), 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(max(ARO_heatmap)/paletteLength, max(ARO_heatmap), length.out=floor(paletteLength/2)))

pheatmap(
  ARO_heatmap,
  annotation = anno,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 9, cellheight = 9, fontsize = 10,
  width = 24, height = 12, annotation_colors = ann_colors, show_colnames = T,
main = "Z-score of ARO gene in Penem class in horses")

fig_name = "sig_changed_ARO_zscore_heatmap_horses_penem.pdf"
print(glue("Figure saved as {fig_name}"))

pheatmap(
  ARO_heatmap,
  annotation = anno,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 12, cellheight = 12, fontsize = 12,
  width = 24, height = 12, annotation_colors = ann_colors, show_colnames = T,
main = "Z-score of ARO gene in Penem class in horses", filename = here("Output_pseudocount", fig_name))

```

### Second heatmap for figure construction

```{r, fig.width=30, fig.height=50, out.width="80%"}

just_sig <- summaries_df %>%
  filter(adj.p.value < 0.05)

my_AROs <- summaries_df %>%
  pivot_wider(id_cols = ARO, names_from = contrast, values_from = adj.p.value) %>%
  filter(ARO %in% just_sig$ARO)

CARD_subset <- CARD_rpkm %>%
  filter(SampleID %in% s_toTest$SampleID) %>%
  filter(ARO %in% my_AROs$ARO)

ARO_z_score <- CARD_subset %>%
  group_by(ARO) %>%
  mutate(ARO_zscore = scale(RPKM)) %>%
  ungroup()

ARO_heatmap <- ARO_z_score %>%
  spread_to_numeric_matrix(row_key = "description", col_key = "SampleID", value = "ARO_zscore")

s_temp <- s_toTest %>%
  column_to_rownames(var = "SampleID")
s_temp <- s_temp[order_for_later,]
ARO_heatmap <- ARO_heatmap[,order_for_later]
colnames(ARO_heatmap) <- s_temp$HorseID

paletteLength <- 50
myColor <- colorRampPalette(c("yellow", "white", "blue"))(paletteLength)
myBreaks <- c(seq(min(AMR_heatmap), 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(max(AMR_heatmap)/paletteLength, max(AMR_heatmap), length.out=floor(paletteLength/2)))

pheatmap(
  ARO_heatmap,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 9, cellheight = 9, fontsize = 10,
  width = 24, height = 12, show_colnames = T,
main = "Z-score of ARO gene in Penem class in horses")

fig_name = "sig_changed_ARO_zscore_heatmap_horses_penem_A.pdf"
print(glue("Figure saved as {fig_name}"))

pheatmap(
  ARO_heatmap,
  color = myColor,
  breaks = myBreaks,
  clustering_distance_rows = "manhattan",
  cluster_cols = F,
  cellwidth = 12, cellheight = 12, fontsize = 12,
  width = 24, height = 12, show_colnames = T,
main = "Z-score of ARO gene in Penem class in horses", filename = here("Output_pseudocount", fig_name))

```


# End of Report

```{r Generate time stamped report2, eval=FALSE, include=FALSE}

#neat: you can run this following command in the console to give your reports custom names (or date-stamp them)

rstudioapi::documentSaveAll()

script_dir <- file.path("Scripts","Rmds")
Output_dir <- file.path("Output_pseudocount")

if (!dir.exists(here::here(Output_dir))) {
  dir.create(here::here(Output_dir))
}

report_fp <- here::here(script_dir, "Shotgun_ARG_analysis_pseudocount.Rmd")
Output_fp <- here::here(Output_dir, paste('CHOPMC-373_Redding_Shotgun_AMR_analysis_with_pseudocount', Sys.Date(), '.pdf', sep=''))

rmarkdown::render(report_fp, output_file = Output_fp)

```
